---
title: "Disentangling length-dependent and length-independent temporal variation in survival and maturation in Atlantic salmon"
authors: "Boulaire et al. (in prep)"
date: "`r Sys.Date()`"
output:
  md_document:
    variant: markdown_github
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

# Disentangling length-dependent and length-independent temporal variation in survival and maturation in Atlantic salmon

## Introduction

### What is this ?

Code and workflow accompanying the manuscript: "Disentangling length-dependent and length-independent temporal variation in survival and maturation in Atlantic salmon".\
The project implements an Integrated Integral Projection Model (IPM²) for the marine phase of the Scorff River Atlantic salmon (*Salmo salar*) population, inferring length-dependent survival and maturation from abundance, molecular-sexing and scale-length data.

------------------------------------------------------------------------

## Overview

### Abstract

Identifying the proximate mechanisms through which environmental changes shape population dynamics is critical to analyse and forecast responses under global change and support ecosystem-based management. Declines in size-at-age have been widely reported across taxa, but understanding how temporal variations in body size at key life stages propagate through the life cycle and shape population dynamics remains challenging, especially for species with partially observable life cycles. Atlantic salmon (*Salmo salar*) have undergone concurrent declines in abundance, length-at-age, and sea age at maturation, largely attributed to deteriorating marine growth during the unobservable oceanic phase. The marine phase of Atlantic salmon is shaped by a complex interplay between survival and maturation schedules, yet the relative contributions of body size to this interplay remain poorly understood. We developed a Bayesian Integrated Integral Projection Model (IPM²) that integrates 24 years (1996-2019) of monitoring data from the Scorff River (France), combining abundance estimates, molecular sexing, and length-at-stage distributions reconstructed from archived fish scales. The model allowed us to disentangle the relative contributions of length-dependent and length-independent processes to temporal variation in survival and maturation, separately for males and females. Length-at-stage explained a substantial part of the pathways to variation in marine vital rates, though considerable length-independent variation remained. Larger individuals consistently experienced higher survival and higher maturation than smaller conspecifics, with females maturing at lower rates than males for the same length. However, temporal variation in the smolt length distribution explained only 9% of the temporal variations in marine survival. In contrast, temporal variation in the post-smolt length distribution explained 45% of the temporal variation in marine maturation for females and 21% for males. The IPM² framework is transferable to other salmon and fish populations with partially observable life cycles to understand how changing growth conditions shape population dynamics under environmental change, and for building climate-enhanced population dynamics and stock assessment models.

### Key words

body size; integral projection model; integrated population model; life-history; *Salmo salar*; temporal variation; vital rates

### Author(s)

***Eliot BOULAIRE***\*\*\
UMR DECOD, Institut Agro, Rennes, France\
pôle MIAME,\
eliot.boulaire\@institut-agro.fr

***Marie NEVOUX***\*\
UMR DECOD, INRAE, Rennes, France\
marie.nevoux\@inrae.fr

***Etienne RIVOT***\*\
UMR DECOD, Institut Agro, Rennes, France\
etienne.rivot\@institut-agro.fr

\*M. NEVOUX & E. RIVOT share senior co-authorship

### Corresponding author

\*\*Comments and requests should be addressed to Eliot BOULAIRE: eliot.boulaire\@institut-agro.fr or eliotboulaire\@gmail.com. All material is free of use, but I would appreciate being told, and this dataet and the matching paper cited if appropriate.

### Acknowledgements

We are grateful to all those involved in the preparation and collection of the data used in this study, including both current and former members of the INRAE UE U3E, the UMR DECOD and the local angling association. We especially thank Frédéric Marchand for overseeing scale data collection as part of the COLISA program, Ludivine Lamireau for conducting the scale readings, and Anne-Laure Besnard, Lisa Meslier, and Thibaut Jousseaume for performing the genetic sexing. Also, a special thanks to Cécile Tréhin for her earlier work on this topic, which laid the groundwork for this publication. We thank Institut Agro, the Brittany Region as well as OFB for funding this study.

------------------------------------------------------------------------

## Data availability & requirements

### Data availability

Raw data are publicly archived and must be placed under `data/rawdata/` (see structure below) before running:

-   Trapping data (counts & biometry): GBIF archive (<https://doi.org/10.15468/yvcw8n>),

-   CMR abundance model (code & results): Zenodo archive (<https://doi.org/10.5281/zenodo.3275148>) & GitHub repository (<https://github.com/ORE-DiaPFC/Abundance>),

-   Scale samples & molecular sexing: COLISA collection (<https://doi.org/10.15454/D3ODJM>),

-   IPM² model (code & results; this repository): French data gouv archive (XXX) & GitHub repository (<https://github.com/eliotboulaire/salmon-marine-IPM2>).

### Software and packages requirement

R version 4.4.2 (or later). Rstudio 2024.12.0.467 (or later).\
MCMC via NIMBLE, which requires a working C++ toolchain (Rtools 4.4 or later on Windows).

| Purpose | Packages |
|------------------------------------|------------------------------------|
| Data wrangling | `dplyr` (1.1.4); `tidyr` (1.3.1); `purrr` (1.0.4) |
| Running model (MCMC) | `nimble` (1.4.0); `parallel` (4.4.1); `coda` (0.19.4.1) |
| Saving models input & output | `qs` (0.27.3) |
| Model diagnostics & comparison | `ggmcmc` (1.5.1.1); `coda` (0.19.4.1) |
| Model comparison | `loo` (2.8.0) |
| Variance partitioning | `relaimpo` (2.2.7) |
| Figures | `ggplot2` (3.5.2); `ggridges` (3.5.2); `ggnewscale` (0.5.2) |

Each script auto-installs any missing packages via a small `pkginstall()` helper at the top of the file.

------------------------------------------------------------------------

## File & folder structure

```{r echo=FALSE}

fs::dir_tree(".", recurse = 3, regexp = "\\.(docx|xlsx|pptx|tif|tmp)$", invert = TRUE)
```

------------------------------------------------------------------------

## File descriptions

*salmon-marine-IPM2.Rproj* - RStudio project file; open this first so all scripts run with the project root as the working directory.

### Manuscript/

*Preprint_Manusrcipt.pdf* - PDF of the preprint. See also:

-   XXX (Link to the preprint),

-   XXX (Link to the final publication).

### Data/

A folder with all the data needed for the article, see Data availability for original archives. Organised in two subfolders `rawdata/` (original and filtered data from archives) and `realdata/` (formatted files for model fitting).

#### Rawdata/

A subfolder organised with original data file and filtered data files per data type and stages `abundances/`, `sexes`/, and `scales/`.

-   *data_Scorff_CMR.csv* - original data file for abundance estimates.

| Columns | Description |
|--------------------|----------------------------------------------------|
| stage | Life-cycle stage: "Smolt" (downstream-migrating juvenile), "1SW" (adult returning after one sea-winter), "2SW" (adult returning after two sea-winters). |
| cohort | Year of smolt migration (smolts) or return (1SW, 2SW); 1996–2019. |
| mean | Mean abundance estimate for the stage and year, from the lognormal CMR model. |
| sd | Standard deviation of the abundance estimate for the stage and year, from the lognormal CMR model. |

-   *data_Scorff_ORE.csv* - original data file for scale length and molecular sexing.

| Columns | Description |
|--------------------|----------------------------------------------------|
| index | Identifier for each individual scale. |
| stage | Life-cycle stage: "Smolt" (downstream-migrating juvenile) or "Adults" (upstream-migrating adults). |
| cohort | Year of smolt migration (smolts) or return (adults); 1996–2019. |
| sexe | Sex of the individual: F (female), M (male), or NA (not sexed). |
| age | Age within stage: "1SW" or "2SW" for adults; "Smolt" for smolts (all freshwater ages pooled). |
| migration | Scale length at seaward migration. Measured retrospectively from the river–sea transition mark on adult scales; for scales sampled from smolts, taken as the scale edge. |
| end1sum | Scale length at the end of the first summer at sea (the maturation "decision" point: mature and return as 1SW, or remain at sea and return as 2SW). Measured retrospectively from adult scales. |

-   *data_create.R* - reads both *data_Scorff* files to build the datasets used by the model (*LogN*, *Ns*, *S*).
-   *data_save.R* - assembles those datasets (*LogN*, *Ns*, *S*) into the model inputs *data.qs* and *const.qs*.

##### abundances/

CMR-estimated abundances in the Scorff River, cohort years 1996–2019:

-   *LogN1.csv* - smolts.

-   *LogN3.csv* - 1SW returns.

-   *LogN4.csv* - 2SW returns.

##### sexes/

Genotyped individual scales in the Scorff River, cohort years 1996–2019:

-   *Ns1.csv* - smolts.

-   *Ns3.csv* - 1SW returns.

-   *Ns4.csv* - 2SW returns.

##### scales/

Individual scale lengths in the Scorff River, cohort years 1996–2019:

-   *S1.csv* - smolts (from smolt scales).

-   *S1surv.csv* - surviving smolts (reconstructed from adult scales).

-   *S2.csv* - post-smolts (reconstructed from adult scales).

-   *S2m.csv* - maturing post-smolts (reconstructed from 1SW returns scales).

-   *S2nm.csv* - non-maturing post-smolts (reconstructed from 2SW returns scales).

#### Realdata/

A subfolder with formatted files for model fitting, organised in model folders `M[0-9]/`.

-   *data.qs* - all data passed to the model.

-   *const.qs* - all constants passed to the model (dimensions, sample sizes, length bounds, fixed survivals).

##### M[0-9]/

-   *monitor1.qs* - monitored parameters retained for results.

-   *monitor2.qs* - monitored parameters retained for PSIS-LOO comparison.

-   *inits_chain[1-6].qs* - initial values for the high-level parameters, on file per MCMC chain.

-   *inits_1chain.qs* - initial values for all parameters, for a single chain (compilation).

-   *inits_nchains.qs* - initial values for all parameters, for all chains (running).

### Functions/

A folder with all function necessary for the project, organised two subfolders `rfunctions/` (all R functions for creating `realdata/` files) and `nimblefunctions/` (all NimbleFunction used in the models).

#### R functions/

-   *f_fillNA.R* - pads the scale-length datasets with *NA* so that all cohorts share the same number of rows despite differing sample sizes.

-   *f_geninit.R* - generates initial values for the high-level parameters to start the MCMC chains.

-   *f_round.R* - round down (floor) or up (ceiling) to set the min and max of the length axis, ensuring all observed lengths fall inside the discretised range.

#### Nimble functions/

-   *nf_l.R* - computes the mean length within each length class, given the mean and sd of the assumed Gaussian length distribution at the population level.

-   *nf_pi.R* - computes the proportion of individuals within each length class, given the mean and sd of the assumed Gaussian length distribution at the population level.

-   *nf_res.R* - back-calculates the mean and sd of the population-level length distribution from the number of individuals and the mean length in each length class.

### Models/

A folder with all model codes, organised in model folders `M[0-9]/`.

#### M[0-9]/

-   *model_code.R* - NIMBLE model code (one variant per model folder).

### Run/

A folder with all the scripts needed to fit the models.

-   *model_save.R* - sets up the monitors and initial values, producing the *monitor.qs* and *inits.qs* files.

-   ***run_model.R*** **- fits the models.**

-   *debug_model.R* - helper for diagnosing problems encountered while fitting.

### Saves/

A folder with all saved MCMC outputs, organised in model folders `M[0-9]/`.

#### M[0-9]/

-   *MCMC.qs* - posterior samples saved at the end of the MCMC run **(not included in the repository -too heavy)**

-   *run_info.txt* - metadata (settings, timing) for the fitted model.

### Graphs/

A folder with all the scripts needed to get the figures and tables outputs from the main manuscript and the supplementary material.

#### Figures/

-   *fig4_structure.R* - R script to recreate Figure 4: Annual variation in length distribution

-   *fig5_size.R* - R script to recreate Figure 5: Annual length-dependent vital rates

-   *fig6_effects.R* - R script to recreate Figure 6: Annual variation in population-level vital rates

-   *figA1_data.R* - R script to recreate figures of Annexe 1 (A1.1; A1.2 & A1.3): Annual data values

-   *figA5_fit.R* - R script to recreate figures of Annexe 5.1 (A5.1.1; A5.1.2 & A5.1.3): Annual estimated posterior distributions against data values

-   *figA7_cor.R* - R script to recreate figures of Annexe 7.1 (A7.1.1 & A7.1.2): Distribution of Pearson correlations between the sex-specific parameters of maturation

-   *figA7_scatter.R* - R script to recreate figures of Annexe 7.2 (A7.2.1 & A7.2.2): Annual variation in length-dependent effect on population-level vital rates

#### Tables/

-   *tab1_loo.R* - R script to recreate Table 1: PSIS-LOO model comparison for the influence of sex, length-independent, and length-dependent effects on vital rates modelling

-   *tabA1_data.R* - R script to recreate tables of Annexe 1 (A1.2 & A1.3): Number of scale samples analysed to characterise sex-ratios and length distributions

-   *tabA4_conv.R* - R script to recreate table of Annexe 4 (A4.1): Convergence diagnosis of each model (Gelman-Rubin-Brooks diagnostic, Geweke diagnostic & ESS)

-   *tabA4_sum.R* - R script to recreate table of Annexe 4 (A4.2): Estimated parameters summaries of each model

-   *tabA5_ppc-abund.R* - R script to recreate table of Annexe 5.2 (A5.2.1): Posterior Predictive Check (PPC) of abundance estimates

-   *tabA5_ppc-sex.R* - R script to recreate table of Annexe 5.2 (A5.2.2): Posterior Predictive Check (PPC) of sex-ratio data

-   *tabA5_ppc-length.R* - R script to recreate table of Annexe 5.2 (A5.2.3; A5.2.4; A5.2.5): Posterior Predictive Check (PPC) of length-distribution data. 3) Proportions; 4) Mean length; 5) 95% IQR

-   *tabA5_ppc-alpha.R* - R script to recreate table of Annexe 5.2 (A5.2.6): Posterior Predictive Check (PPC) of vital rates alpha parameters estimates

### Results/

#### Figures/

-   *fig1.pdf* - Figure 1:Stage-structured model of the marine phase of the salmon life cycle, with time steps indicated.
-   *fig2.pdf* - Figure 2: Photographs of salmon scales from a smolt (left), a 1SW (middle), and a 2SW (right), each sampled from a different individual, showing the smolt mark (green line) and the post-smolt mark (purple line).
-   *fig3.pdf* - Figure 3: Schematic representation of the inference process of length-dependent vital rates.
-   *fig4.pdf* - Figure 4: Annual variations in length distributions estimated by the best-supported model (M1).
-   *fig5.pdf* - Figure 5: Annual length-dependent survival and maturation rates estimated from the best-supported model (M1).
-   *fig6.pdf* - Figure 6: Annual variation in population-level survival and maturation rates, integrated across all length classes, with their length-dependent and length-independent components, estimated from the best-supported model (M1).
-   *figA1.1.pdf* - Figure A1.1: Annual abundance of smolts, 1SW and 2SW adults in the River Scorff estimated from the CMR model, by year of smolt migration (1996-2019).
-   *figA1.2.pdf* - Figure A1.2: Annual proportion of females in the sample of genotyped scales in the River Scorff, by year of smolt migration (1996-2019).
-   *figA1.3.pdf* - Figure A1.3: Annual length distributions derived from scale samples for smolts, surviving smolts, post-smolts, maturing post-smolts and non-maturing post-smolts in the River Scorff, by year of smolt migration (1996-2019).

##### M[0-9]/

-   *figA5.1.1.pdf* - Figure A5.1.1: Annual abundance of smolts, 1SW and 2SW in the River Scorff estimated from the CMR model (red) and from the IPM² model (black), by year of smolt migration (1996-2019).
-   *figA5.1.2.pdf* - Figure A5.1.2: Annual proportion of females among genotyped scales (red) and as estimated from the IPM² model (black), for smolts, 1SW and 2SW in the River Scorff, by year of smolt migration (1996-2019).
-   *figA5.1.3.pdf* - Figure A5.1.3: Annual length distributions derived from scale samples (red) and as estimated from the IPM² model (black), for smolts, surviving smolts, post-smolts, maturing post-smolts and non-maturing post-smolts in the River Scorff, by year of smolt migration (1996-2019).

#### Tables/

-   *tab1.csv* - Table 1: PSIS-LOO model comparison for the influence of sex, length-independent, and length-dependent effects of survival and maturation modelling.
-   *tabA1.2.csv* - Table A1.2: Number of scale samples used to inform sex ratio, by stage (smolt, 1SW, 2SW) and year of smolt migration (1996-2019).
-   *tabA1.3.csv* - Table A1.3 Number of scale samples analysed to characterise length distributions, by stage (smolt, 1SW, 2SW) and year of smolt migration (1996-2019). One scale per fish was analysed.

##### M[0-9]/

-   *tabA4.1.csv* - Table A4.1: Convergence diagnosis of each IPM² model (Gelman-Rubin-Brooks diagnostic, Geweke diagnostic & ESS diagnostic).

-   tabA4.2.csv *-* Table A4.2: Estimated parameters summaries of each IPM² model.

##### M1/

-   *tabA5.2.1.csv* - Table A5.2.1: Bayesian p-values quantifying discrepancy between observed and replicated abundances by stage (smolt, 1SW, 2SW) and year of smolt migration (1996-2019).

-   *tabA5.2.2.csv* - Table A5.2.2: Bayesian p-values quantifying discrepancy between observed and replicated sex ratios by stage (smolt, 1SW, 2SW) and year of smolt migration (1996-2019).

-   *tabA5.2.3.csv* - Table A5.2.3: Bayesian p-values quantifying discrepancy between observed and replicated proportion of individuals per length class by stage (smolts, surviving smolts, post-smolts, maturing post-smolts and non-maturing post-smolts) and year of smolt migration (199

-   *tabA5.2.4.csv* - Table A5.2.4: Bayesian p-values quantifying discrepancy between observed and replicated mean length by stage (smolts, surviving smolts, post-smolts, maturing post-smolts and non-maturing post-smolts) and year of smolt migration (1996-2019).

-   *tabA5.2.5.csv* - Table A5.2.5: Bayesian p-values quantifying discrepancy between observed and replicated 95% quantile ranges of length by stage (smolts, surviving smolts, post-smolts, maturing post-smolts and non-maturing post-smolts) and year of smolt migration (1996-2019).

-   *tabA5.2.6A.csv* - Table A5.2.6A: Bayesian p-values testing the normality (*χ²*) and independence across years of smolt migration (*AR(1)*) of the length-independent random terms of the survival rate.

-   *tabA5.2.6B.csv* - Table A5.2.6B: Bayesian p-values testing the normality (*χ²*) and independence across years of smolt migration (*AR(1)*) of the length-independent random terms of the maturation rate.

------------------------------------------------------------------------

## Execution instructions

Open *MODESTA.Rproj* first, then run the scripts in the following order. Raw data must already be in `data/rawdata/` (see Data availability).

1.  **Prepare the data** - run *data_create.R*, then *data_save.R*. Optionally check the inputs with *figA1_data.R* and *tabA1_data.R*.

2.  **Set up the run** - run *model_save.R* to generate the monitors and initial values.

3.  **Fit the models** - run *run_model.R*. If problems occur, re-check the previous steps and use *debug_model.R*.

4.  **Produce figures and tables** - once *MCMC.qs* and *run_info.txt* exist, run the post-processing scripts. Recommended order:

    1.  **MCMC diagnostics -** *tabA4_conv.R*, *tabA4_sum.R*

    2.  **Model comparison -** *tab1_loo.R*

    3.  **Model evaluation** - *figA5_fit.R* & *tabA5_ppc-abund.R*; *tabA5_ppc-sex.R*; *tabA5_ppc-length.R*; *tabA5_ppc-alpha.R*

    4.  **Main article figures** - *fig4_structure.R*; *fig5_size.R* & *fig6_effects.R*

    5.  **Supplementary material figures** -*figA7_cor.R* & *figA7_scatter.R*

------------------------------------------------------------------------

## License

The content of this project itself is licensed under the [GNU GPL v3.0 license](https://www.gnu.org/licenses/gpl-3.0.en.html), except the dataets which are under the [Creative Commons Attribution 4.0 license](https://creativecommons.org/licenses/by/4.0/).
